Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-12-02 manu_draft.Rmd
adding some references to discussion

The more recently developed method of weighted regression on time, discharge, and season (WRTDS) uses a more general local regression scheme than STL [@Hirsch10;@Beck15].  Designed for river data where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allows parameters to vary smoothly in relation to both time and discharge.  This yields parameters that are specific to season, year, and flow regime.  Conceptually, the approach is similar to localized multi-polynomial smoothing methods, although the application was developed specifically for describing long-term water quality trends.  Standard error estimates of predictions are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15]. Although a useful addition to the original method in @Hirsch10, the approach requires extensive resampling as a post-hoc application to a previously fitted model.

@@

The approach developed here can be used to estimate seasonal averages with uncertainty in water quality parameters and evaluate trends in these seasonal changes across years using full propagation of error estimates across methods.  This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv package that uses generalized cross-validation to extract an underlying signal in the observed data. Importantly, specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.  A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risk inflated Type I error rates. We further show that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package.  This comparison demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time. Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses to estimate significant of trends developed herein.

The use of the GAM results with second and third stage analyses to assess trends is a new approach that previous trend analysis methods cannot sufficiently address.  Although the development of our approach was motivated by specific questions regarding assessment of seasonal changes over time, the application has advantages over more conventional methods that inadequately account for time series characteristics of water quality data from long-term monitoring programs. In particular, missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design.  As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data.  These methods only assess the direction and significance of comparisons between year pairs, and importantly, do not account for full propagation of uncertainty inherent in raw observations.  Aggregation of raw data, e.g., averaging of observations within a year or season to comply with the requirements of Kendall tests, risks loss of information by removing variation between observations at smaller time scales. The logical outcome is increased likelihood of incorrect conclusions from test results.

Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.  Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  This is particularly problematic if sampling effort varies substantially between years.  Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only).  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages. More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses.  As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.

>>>>

The more recently developed method of weighted regression on time, discharge, and season (WRTDS) uses a more general local regression scheme than STL [@Hirsch10;@Beck15;@Beck18b].  Designed for river data where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allows parameters to vary smoothly in relation to both time and discharge.  This yields parameters that are specific to season, year, and flow regime.  Conceptually, the approach is similar to localized multi-polynomial smoothing methods, although the application was developed specifically for describing long-term water quality trends.  Standard error estimates of predictions are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15]. Although a useful addition to the original method in @Hirsch10, the approach requires extensive resampling as a post-hoc application to a previously fitted model.

@@

The approach developed here can be used to estimate seasonal averages with uncertainty in water quality parameters and evaluate trends in these seasonal changes across years using full propagation of error estimates across methods.  This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv R package that uses generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17]. Importantly, specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.  A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risk inflated Type I error rates. We further demonstrated that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).  This comparison demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time. Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses to estimate significant of trends developed herein.

The use of the GAM results with second and third stage analyses to assess trends is a new approach that previous trend analysis methods cannot sufficiently address.  Although the development of our approach was motivated by specific questions regarding assessment of seasonal changes over time, the application has advantages over more conventional methods that inadequately account for time series characteristics of water quality data from long-term monitoring programs. In particular, missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design [@Junninen04;@Racault14].  As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data [@Hirsch82;@Helsel20].  These methods only assess the direction and significance of comparisons between year pairs, and importantly, do not account for full propagation of uncertainty inherent in raw observations.  Aggregation of raw data, e.g., averaging of observations within a year or season to comply with the requirements of Kendall tests, risks loss of information by removing variation between observations at smaller time scales. The logical outcome is increased likelihood of incorrect conclusions from test results.

Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.  Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  This is particularly problematic if sampling effort varies substantially between years.  Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only)[@Fouquet12].  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages. More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses.  As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.
